---
layout: page
title: Programming with R
subtitle: Analyzing raptor data
minutes: 30
---


> ## Learning Objectives {.objectives}
> * Exposure to the seven core programming concepts
> * Read tabular data from a file into a program.
> * Assign values to variables.
> * Select individual values and subsections from data.
> * Perform operations on a data frame of data.
> * Display simple graphs.


### The Seven Core Concepts


As noted by Greg Wilson (the founder of Software Carpentry), every
programming language shares [seven core elements][1]:

1.  Individual things (the number 2, the character 'hello')
2.  Commands that operate on things (the + symbol, the `length`
    function)
3.  Groups of things (R vectors, dataframes, and arrays)
4.  Ways to repeat yourself (for and while loops)
5.  Ways to create chunks (functions, objects/classes, and packages)
6.  Ways to make choices (if and try statements)
7.  Ways to combine chunks (function composition)

The lines between these are often blurry in practice - they are purely
a conceptual framework that helps programmers write code that does
what they want it to do.

We expect that you'll find the basics of 1 and 2 fairly
straightforward. We'll go quickly through those and will spend the
most time on items 3-6. We won't really talk about 7, as it is not as
common in R programming as it is in, say, shell scripting (pipes and
redirection).

Don't worry if you don't already know what all of the above examples
mean - you'll know by the end of this lesson.


### Starting with RStudio


To open RStudio, click on the RStudio icon in the Applications (if you
are on a Mac) or in your Programs if you are on Windows.

There are four windows in RStudio that we will refer to throughout the
workshop

1. *The R Script*: Typically the upper left hand corner of RSutdio.
   This is where you write R code that you can save and reuse later.
2. *The R Console*: Typically in the lower left hand corner. This is
   where you execute R code.
3. *The R Environment*: Typically in the upper right hand corner. This
   is where you can see defined variables.
4. *R "Information"*: Typically in the lower right hand corner. This
   is where you see plots, help and other R information.

For all the work in this workshop, we will be typing code in the R
script and then executing it in the R console.  For simple commands
(e.g. `2 + 2`) this may seem stupid, but writing in a script will help
all of your work be reproducible! Think of the script as your lab
notebook.

> **TIP**: Some helpful R studio shortcuts
>
> 1. Run the current line of selection
>   - Windows: `Ctrl-Enter`
>   - Mac: `Command-Enter`
> 2. Source the entire script
>   - Windows: `Ctrl-Shift-Enter`
>   - Mac: `Command-Shift-Enter`


### Arthritic raptors


We are studying inflammation in raptors who have been given a new
treatment for injuries, and need to analyze the first dozen data sets.
The data sets are stored in
[comma-separated values](reference.html#comma-separated-values-(csv))
(CSV) format. Each row holds the observations for just one
raptor. Each column holds the inflammation measured in a day, so we
have a set of values in successive days.  The first few rows of our
first file look like this:


~~~{.output}
0,0,1,3,1,2,4,7,8,3,3,3,10,5,7,4,7,7,12,18,6,13,11,11,7,7,4,6,8,8,4,4,5,7,3,4,2,3,0,0
0,1,2,1,2,1,3,2,2,6,10,11,5,9,4,4,7,16,8,6,18,4,12,5,12,7,11,5,11,3,3,5,4,4,5,5,1,1,0,1
0,1,1,3,3,2,6,2,5,9,5,7,4,5,4,15,5,11,9,10,19,14,12,17,7,12,11,7,4,2,10,5,4,2,2,3,2,2,1,1
0,0,2,0,4,2,2,1,6,7,10,7,9,13,8,8,15,10,10,7,17,4,4,7,6,15,6,4,9,11,3,5,6,3,3,4,2,3,2,1
0,1,1,3,3,1,3,5,2,4,4,7,6,5,3,10,8,10,6,17,9,14,9,7,13,9,12,6,7,7,9,6,3,2,2,4,2,0,1,1

~~~

We want to:

* Load data into memory,
* Calculate the average value of inflammation per day across all
  raptors, and
* Plot the results.

To do all that, we'll have to learn a little bit about programming.

### Loading Data

To load our inflammation data, first we need to tell our computer
where is the file that contains the values. We have been told its name
is `inflammation-01.csv`. This is very important in R, if we forget
this step we’ll get an error message when trying to read the file. We
can change the current working directory using the function
`setwd`. For this example, we change the path to the directory we just
created:

```{r,eval=FALSE}
setwd("~/Desktop/2017-01-12-ucb/lessons/R")
```

Just like in the Unix Shell, we type the command and then press
`Enter` (or `return`).  Alternatively you can change the working
directory using the RStudio GUI using the menu option `Session` ->
`Set Working Directory` -> `Choose Directory...`

The data files are located in the directory `data` inside the working
directory. Now we can load the data into R using `read.csv`:

```{r, results="hide"}
read.csv(file = "data/inflammation-01.csv", header = FALSE)
```

The expression `read.csv(...)` is a
[function call](reference.html#function-call) that asks R to run the
function `read.csv`.

`read.csv` has two [arguments](reference.html#argument): the name of
the file we want to read, and whether the first line of the file
contains names for the columns of data.  The filename needs to be a
character string (or [string](reference.html#string) for short), so we
put it in quotes. Assigning the second argument, `header`, to be
`FALSE` indicates that the data file does not have column
headers. We'll talk more about the value `FALSE`, and its converse
`TRUE`, in lesson 04.

> ## Tip {.callout}
>
> `read.csv` actually has many more arguments that you may find useful
> when
> importing your own data in the future. You can learn more about these
> options in this supplementary [lesson](01-supp-read-write-csv.html).

The utility of a function is that it will perform its given action on
whatever value is passed to the named argument(s).  For example, in
this case if we provided the name of a different file to the argument
`file`, `read.csv` would read it instead.  We'll learn more of the
details about functions and their arguments in the next lesson.

Since we didn't tell it to do anything else with the function's
output, the console will display the full contents of the file
`inflammation-01.csv`.  Try it out.

`read.csv` read the file, but we can't use data unless we assign it to
a variable.  A variable is just a name for a value, such as `x`,
`current_temperature`, or `subject_id`.  We can create a new variable
simply by assigning a value to it using `<-`

```{r}
weight_kg <- 55
```

Once a variable has a value, we can print it by typing the name of the
variable and hitting `Enter` (or `return`).  In general, R will print
to the console any object returned by a function or operation *unless*
we assign it to a variable.

```{r}
weight_kg
```

We can do arithmetic with the variable:

```{r}
# weight in pounds:
2.2 * weight_kg
```

> ## Tip {.callout}
> We can add comments to our code using the `#`
>character. It is useful to document our code in this way so that
>others (and us the next time we read it) have an easier time
>following what the code is doing.

We can also change an object's value by assigning it a new value:

```{r}
weight_kg <- 57.5
# weight in kilograms is now
weight_kg
```

If we imagine the variable as a sticky note with a name written on it,
assignment is like putting the sticky note on a particular value:

<img src="fig/python-sticky-note-variables-01.svg" alt="Variables as
Sticky Notes" />

This means that assigning a value to one object does not change the
values of other variables. For example, let's store the subject's
weight in pounds in a variable:

```{r}
weight_lb <- 2.2 * weight_kg
# weight in kg...
weight_kg
# ...and in pounds
weight_lb
```

<img src="fig/python-sticky-note-variables-02.svg" alt="Creating
Another Variable" />

and then change `weight_kg`:

```{r}
weight_kg <- 100.0
# weight in kg now...
weight_kg
# ...and weight in pounds still
weight_lb
```

<img src="fig/python-sticky-note-variables-03.svg" alt="Updating a
Variable" />

Since `weight_lb` doesn't "remember" where its value came from, it
isn't automatically updated when `weight_kg` changes.  This is
different from the way spreadsheets work.

> ## Tip {.callout}
> An alternative way to print the value of a
>variable is to use () around the assignment statement. As an example:
>`(total_weight <- weight_kg + weight_lb)`, adds the values of
>`weight_kg` and `weight_lb`, assigns the result to the
>`total_weight`, and finally prints the assigned value of the variable
>`total_weight`.


Now that we know how to assign things to variables, let's re-run
`read.csv` and save its result:

```{r}
dat <- read.csv(file = "data/inflammation-01.csv", header = FALSE)
```

This statement doesn't produce any output because assignment doesn't
display anything.  If we want to check that our data has been loaded,
we can print the variable's value.  However, for large data sets it is
convenient to use the function `head` to display only the first few
rows of data.

```{r}
head(dat)
```

> ## Challenge - Assigning values to variables {.challenge}
>
> Draw diagrams showing what variables refer to what values after each
>statement in the following program:
>
~~~{.r}
mass <- 47.5
age <- 122
mass <- mass * 2.0
age <- age - 20
~~~

### Manipulating Data

Now that our data is loaded in memory, we can start doing things with it.
First, let's ask what type of thing `dat` is:

```{r}
class(dat)
```

The output tells us that is a data frame. Think of this structure as a
spreadsheet in MS Excel that many of us are familiar with.  Data
frames are very useful for storing data and you will find them
elsewhere when programming in R. A typical data frame of experimental
data contains individual observations in rows and variables in
columns.

We can see the shape, or
[dimensions](reference.html#dimensions-(of-an-array)), of the data
frame with the function `dim`:

```{r}
dim(dat)
```

This tells us that our data frame, `dat`, has `r nrow(dat)` rows and
`r ncol(dat)` columns.

If we want to get a single value from the data frame, we can provide
an [index](reference.html#index) in square brackets, just as we do in
math:

```{r}
# first value in dat
dat[1, 1]
# middle value in dat
dat[30, 20]
```

An index like `[30, 20]` selects a single element of a data frame, but
we can select whole sections as well.  For example, we can select the
first ten days (columns) of values for the first four raptors (rows)
like this:

```{r}
dat[1:4, 1:10]
```

The [slice](reference.html#slice) `1:4` means, "Start at index 1 and
go to index 4."

The slice does not need to start at 1, e.g. the line below selects
rows 5 through 10:

```{r}
dat[5:10, 1:10]
```
We can use the function `c`, which stands for **c**ombine, to select
non-contiguous values:

```{r}
dat[c(3, 8, 37, 56), c(10, 14, 29)]
```

We also don't have to provide a slice for either the rows or the
columns.  If we don't include a slice for the rows, R returns all the
rows; if we don't include a slice for the columns, R returns all the
columns.  If we don't provide a slice for either rows or columns,
e.g. `dat[, ]`, R returns the full data frame.

```{r}
# All columns from row 5
dat[5, ]
# All rows from column 16
dat[, 16]
```

Now let's perform some common mathematical operations to learn about
our inflammation data.  When analyzing data we often want to look at
partial statistics, such as the maximum value per raptor or the
average value per day.  One way to do this is to select the data we
want to create a new temporary data frame, and then perform the
calculation on this subset:

```{r}
# first row, all of the columns
raptor_1 <- dat[1, ]
# max inflammation for raptor 1
max(raptor_1)
```

We don't actually need to store the row in a variable of its own.
Instead, we can combine the selection and the function call:

```{r}
# max inflammation for raptor 2
max(dat[2, ])
```

R also has functions for other common calculations, e.g. finding the
minimum, mean, median, and standard deviation of the data:

```{r}
# minimum inflammation on day 7
min(dat[, 7])
# mean inflammation on day 7
mean(dat[, 7])
# median inflammation on day 7
median(dat[, 7])
# standard deviation of inflammation on day 7
sd(dat[, 7])
```

What if we need the maximum inflammation for all raptors, or the
average for each day?

As the diagram below shows, we want to perform the operation across a
margin of the data frame:

<img src="fig/r-operations-across-axes.svg" alt="Operations Across Axes" />

To support this, we can use the `apply` function.

> ## Tip {.callout}
>
> To learn about a function in R, e.g. `apply`, we can read its help
> documention by running `help(apply)` or `?apply`.

`apply` allows us to repeat a function on all of the rows (`MARGIN =
1`) or columns (`MARGIN = 2`) of a data frame.

Thus, to obtain the average inflammation of each raptor we will need
to calculate the mean of all of the rows (`MARGIN = 1`) of the data
frame.

```{r}
avg_raptor_inflammation <- apply(X=dat, MARGIN=1, FUN=mean)
```

And to obtain the average inflammation of each day we will need to
calculate the mean of all of the columns (`MARGIN = 2`) of the data
frame.

```{r}
avg_day_inflammation <- apply(dat, 2, mean)
```

Since the second argument to `apply` is `MARGIN`, the above command is
equivalent to `apply(dat, MARGIN = 2, mean)`.  We'll learn why this is
so in the next lesson.

> ## Tip {.callout}
>
> Some common operations have more efficient alternatives. For
> example, you can calculate the row-wise or column-wise means with
> `rowMeans` and `colMeans`, respectively.
>
> ## Challenge - Slicing (subsetting) data {.challenge}
>
> A subsection of a data frame is called a
> [slice](reference.html#slice).  We can take slices of character
> vectors as well:
>
> ```{r}
> animal <- c("m", "o", "n", "k", "e", "y")
> # first three characters
> animal[1:3]
> # last three characters
> animal[4:6]
> ```
>
> 1.  If the first four characters are selected using the slice
>`animal[1:4]`, how can we obtain the first four characters in reverse
>order?
>
> 1.  What is `animal[-1]`?
>    What is `animal[-4]`?
>    Given those answers,
>    explain what `animal[-1:-4]` does.
>
> 1.  Use a slice of `animal` to create a new character vector that spells the word "eon", i.e. `c("e", "o", "n")`.


> ## Challenge - Subsetting data 2 {.challenge}
>
> Suppose you want to determine the maximum inflamation for raptor 5 across days three to seven.
> To do this you would extract the relevant slice from the data frame and calculate the maximum value.
> Which of the following lines of R code gives the correct answer?
>
> (a) `max(dat[5, ])`
> (b) `max(dat[3:7, 5])`
> (c) `max(dat[5, 3:7])`
> (d) `max(dat[5, 3, 7])`

### Plotting

The mathematician Richard Hamming once said, "The purpose of computing
is insight, not numbers," and the best way to develop insight is often
to visualize data.  Visualization deserves an entire lecture (or
course) of its own, but we can explore a few of R's plotting features.

Let's take a look at the average inflammation over time.  Recall that
we already calculated these values above using `apply(dat, 2, mean)`
and saved them in the variable `avg_day_inflammation`.  Plotting the
values is done with the function `plot`.

```{r plot-avg-inflammation}
plot(avg_day_inflammation)
```

Above, we gave the function `plot` a vector of numbers corresponding
to the average inflammation per day across all raptors.  `plot`
created a scatter plot where the y-axis is the average inflammation
level and the x-axis is the order, or index, of the values in the
vector, which in this case correspond to the 40 days of treatment.
The result is roughly a linear rise and fall, which is suspicious:
based on other studies, we expect a sharper rise and slower fall.
Let's have a look at two other statistics: the maximum and minimum
inflammation per day.

```{r plot-max-inflammation}
max_day_inflammation <- apply(dat, 2, max)
plot(max_day_inflammation)
```

```{r plot-min-inflammation}
min_day_inflammation <- apply(dat, 2, min)
plot(min_day_inflammation)
```

The maximum value rises and falls perfectly smoothly, while the
minimum seems to be a step function. Neither result seems particularly
likely, so either there's a mistake in our calculations or something
is wrong with our data.

> ## Challenge - Plotting data {.challenge}
>
> Create a plot showing the standard deviation of the inflammation data for each day across all raptors.

> ## Key Points {.callout}
>
> * Use `variable <- value` to assign a value to a variable in order to record it in memory.
> * Objects are created on demand whenever a value is assigned to them.
> * The function `dim` gives the dimensions of a data frame.
> * Use `object[x, y]` to select a single element from a data frame.
> * Use `from:to` to specify a sequence that includes the indices from `from` to
>`to`.
> * All the indexing and slicing that works on data frames also works on vectors.
> * Use `#` to add comments to programs.
> * Use `mean`, `max`, `min` and `sd` to calculate simple statistics.
> * Use `apply` to calculate statistics across the rows or columns of a data frame.
> * Use `plot` to create simple visualizations.

> ## Next Steps {.callout}
>
> Our work so far has convinced us that something's wrong with our first data file.
> We would like to check the other 11 the same way, but typing in the same commands repeatedly is tedious and error-prone.
> Since computers don't get bored (that we know of), we should create a way to do a complete analysis with a single command, and then figure out how to repeat that step once for each file.
> These operations are the subjects of the next two lessons.
