---
layout: page
title: Programming with R
subtitle: Debugging
minutes: 30
---


> ## Learning Objectives {.objectives}
>
> * Exposure to different debugging strategies and options in R
> * Explain strategies to minimize time debugging
> * Debug a function

When I am embarking upon a particular analysis, I first outline the
different functions I will need, their input and outputs. I then write
the functions in the order in which they are needed, testing/debugging
the functions along the way.

### Basic debugging strategies
(modified by a tutorial from Chris Paciorek)

- Read and think about the error message. Sometimes it's inscrutable,
but often it just needs a bit of deciphering. Looking up a given error
message in the
[R mailing list archive](http://tolstoy.newcastle.edu.au/R) or on
Stack Overflow or simply doing a web search with the exact message in
double quotes can be a good strategy.

- Fix errors from the top down --- fix the first error that is
reported, because later errors are often caused by the initial
error. It's common to have a string of many errors, which looks
daunting, caused by a single initial error.

- Is the bug reproducible -- does it always happen in the same way at
at the same point? It can help to restart R and see if the bug
persists -- this can sometimes help in figuring out if there is a
scoping issue (your function is not looking for objects where you
think it is) and we are using a global variable that we did not mean
to.

- Another basic strategy is to build up code in pieces (or tear it
back in pieces to a simpler version). This allows you to isolate where
the error is occurring.

- If you've written your code modularly (functions are not nested
within each other) with lots of functions, you can test individual
functions. Often the error will be in what gets passed into and out of
each function.

- At the beginning of time (the 1970s?), the standard debugging
strategy was to insert print statements in one's code to see the value
of a variable and thereby decipher what could be going wrong. We have
better tools nowadays.

- R is a scripting language, so you can usually run your code line by
line to figure out what is happening. This can be a decent approach,
particularly for simple code. However, when you are trying to find
errors that occur within a series of many nested function calls or
when the errors involve variable scoping (how R looks for variables
that are not local to a function), or in other complicated situations,
using formal debugging tools can be much more effective.  Finally, if
the error occurs inside of functions provided by R, rather than ones
you write, it can be hard to run the code in those functions line by
line.


## Some common causes of bugs

Some of these are R-specific, while others are common to a variety of
languages.

 - Parenthesis mis-matches
 - `[[...]]` vs. `[...]`
 - `==` vs. `=`
 - Silent type conversion when you don't want it, or lack of coercion
   where you're expecting it!!!!!!
 - Using the wrong function or variable name
 - Giving unnamed arguments to a function in the wrong order
 - Forgetting to define a variable in the environment of a function
   and having R, get that variable as a global variable from one of
   the enclosing environments. At best the types are not compatible
   and you get an error; at worst, you use a garbage value and the bug
   is hard to trace. In some cases your code may work fine when you
   develop the code (if the variable exists in the enclosing
   environment), but then may not work when you restart R if the
   variable no longer exists or is different.

More insidious bugs
 - Comparing real numbers exactly using `==` is dangerous because
   numbers on a computer are only represented to limited numerical
   precision. For example,

```{r}
    1/3 == 4*(4/12-3/12)
```
  - You want to compare an entire vector but your code just compares
   the first value (e.g., in an if statement) -- consider using
   *identical* or *all.equal*
  - In an if-else statement, the `else` cannot be on its own line
   (unless all the code is enclosed in `{}`) because R will see the
   `if` part of the statement, which is a valid R statement, will
   execute that, and then will encounter the `else` and return an
   error.

## General tips to avoid bugs

 - Use core R functionality and algorithms already coded. Figure out
   if a functionality already exists in (or can be adapted from) an R
   package (or potentially in a C/Fortran library/package): code that
   is part of standard mathematical/numerical packages will probably
   be more efficient and bug-free than anything you would write.
 - Code in a modular fashion, making good use of functions, so that
   you don't need to debug the same code multiple times. Smaller
   functions are easier to debug, easier to understand, and can be
   combined in a modular fashion (like the UNIX utilities).
 - Write code for clarity and accuracy first; then worry about
   efficiency. Write an initial version of the code in the simplest
   way, without trying to be efficient (e.g., you might use for loops
   even if you're coding in R); then make a second version that
   employs efficiency tricks and check that both produce the same
   output.
 - Plan out your code in advance, including all special
   cases/possibilities.
 - Write tests for your code early in the process.
 - Build up code in pieces, testing along the way. Make big changes in
   small steps, sequentially checking to see if the code has broken on
   test case(s).
 - Don't hard code numbers
 - use variables (e.g., number of
   iterations, parameter values in simulations), even if you don't
   expect to change the value, as this makes the code more readable
   and reduces bugs when you use the same number multiple times;
   e.g. `speedOfLight <- 3e8` or `nIts <- 1000`.


#### R options for debigging

## Interactive debugging via the browser

The core strategy for interactive debugging is to use the *browser*
function, which pauses the current execution, and provides an
interpreter, allowing you to view the current state of R. You can
invoke *browser* in four ways

 - by inserting a call to `browser()` in your code if you suspect
   where things are going wrong

 - by invoking the browser after every step of a function using
   *debug*

 - by using `options(error = recover)` to invoke the browser when an
   error occurs

 - by temporarily modifying a function to allow browsing using *trace*

Once in the browser, you can execute any R commands you want. In
particular, using *ls* to look at the objects residing in the current
function environment, looking at the values of objects, and examining
the classes of objects is often helpful.

## Using *debug* to step through code

To step through a function, use `debug(nameOfFunction)`. Then run your
code. When the function is executed, R will pause execution just
before the first line of the function. You are now using the browser
and can examine the state of R and execute R statements.

Once in the browser context, you can use 'n' or return to step to the
next line, 'c' to execute the entire current function or current loop,
and `' to stop debugging. We'll see this in the screencast demo.

To unflag the function so that calling it doesn't invoke debug, use
`undebug(nameOfFunction)`. In addition to working with functions you
write you can use *debug* with standard R functions and functions from
packages. For example you could do `debug(glm)`.

## Tracing errors in the call stack

*traceback* and *recover* allow you to see the call stack at the time
 of the error - i.e., they will show you all the functions that have
 been called, in the order called. This helps pinpoint where in a
 series of function calls the error may be occurring.

If you've run the code and gotten an error, you can invoke *traceback*
after things have gone awry. R will show you the call stack, which can
help pinpoint where an error is occurring.

More helpful is to be able to browse within the call stack. To do this
invoke `options(error = recover)` (potentially in your *.Rprofile* if
you do a lot of programming). Then when an error occurs, *recover*
gets called, usually from the function in which the error
occurred. The call to *recover* allows you to navigate the stack of
active function calls at the time of the error and browse within the
desired call. You just enter the number of the call you'd like to
enter (or 0 to exit). You can then look around in the frame of a given
function, entering an empty line when you want to return to the list
of calls again.

You can also combine this with `options(warn = 2)`, which turns
warnings into errors to get to the point where a warning was issued.

## Using *trace* to temporarily insert code

*trace* lets you temporarily insert code into a function (including
 standard R functions and functions in packages!) that can then be
 easily removed. You can use trace in a variety of ways.

The most flexible way to use *trace* is to use the argument `edit =
TRUE` and then insert whatever code you want wherever you want in the
function given as the first argument to *trace*. If I want to ensure I
use a particular editor, such as emacs, I can use the argument `edit =
“emacs”`. A standard approach would be to add a line with `browser()`
at some point in the function to be able to step through the code from
that point.

You can also use *trace* without directly editing the function. Here
are a couple examples:

 - `trace(lm, recover)` # invoke *recover* when the function (*lm* in
   this case) starts
 - `trace(lm, exit = browser)` # invoke *browser* when the function
   ends

You call *untrace*, e.g., `untrace(lm)`, to remove the temporarily
inserted code; otherwise it's removed when the session ends.

To figure out why warnings are being issued, you can do
`trace(warning, recover)` which will insert a call to *recover*
whenever *warning* is called.

Of course you can manually change the code in a function without using
*trace*, but it's very easy to forget to change things back (and a
pain to remember exactly what you changed) and hard to do this with
functions in packages, so *trace* is a nice way to do things.

## Challenge - Debugging {.challenge}

Use browser() to find the bug in this function to plot the number of
bee operations in New Zealand

```{r, eval=FALSE}
 hb <- read.csv('data/beeOperations.csv')
 hb <- hb[order(hb$year),]

 plotVar <- function(hb, var, ylabs,
                    types=c("hobby", "semicom", "com"),
                    div=10^3){
    ## function to plot honey bee data by region
    ## takes hb, a data.frame with the varaibles to be plotted
    ## var, the name of the column of interest
    ## types, the different types of honey bee keepers
    ## div, a scalar to divide the values on the yaxis by to avoid
    ## ugly numbers
    layout(matrix(1:(length(unique(hb$region))+1), ncol=2))
    par(oma=c(3,3,2,1), mar=c(2,2,3,1), mgp=c(2,1,0),
        cex.axis=1.5)
    ntypes <- length(types)
    for(i in unique(hb$region)){
       this.region <- hb[hb$region == i,]
        plot(x=this.region$year[this.region$type == types[1]],
             y=this.region[,var][this.region$type == types[1]],
             type="l",
             ylim=c(0, max(hb[,var], na.rm=TRUE)),
             xlab="Year",
             ylab=ylabs,
             yaxt="n",
             main=i)
        ltys <- c("dashed", "dotted")
        for(j in 2:length(types)){
        points(x=this.region$year[this.region$type == types[j]],
               y=this.region[,var][this.region$type == types[j]],
               type="l", lty=ltys[j])
        }
        if(i %in% unique(hb$region)[1:4]){
            axis(2, pretty(c(0, max(hb[,var], na.rm=TRUE))),
                 labels=pretty(c(0, max(hb[,var], na.rm=TRUE)))/div)
        }
    }
}


plotVar(hb, "hives", "Hives")
plotVar(hb, "apiaries", "Apiaries")
plotVar(hb, "beekeepers", "Beekeepers")
plotVar(hb, "hives_beekeeper", "Hives per beekeeper")
plotVar(hb, "apiary_beekeeper", "Apiaries per beekeeper", div=1)
```



##  Practice defensive programming

When writing functions, and software more generally, you'll want to
warn the user or stop execution when there is an error and exit
gracefully, giving the user some idea of what happened. Here are some
things to consider:

 - check function inputs and warn users if the code will do something
   they might not expect or makes particular choices;
 - check inputs to `if` and the ranges in `for` loops;
 - provide reasonable default arguments;
 - document the range of valid inputs;
 - check that the output produced is valid; and
 - stop execution based on checks and give an informative error
   message.

The *warning* and *stop* functions allow you to do stop execution and
issue warnings or errors in the same way that base R code does; in
general they would be called based on an `if` statement. More
succinctly, to stop code if a condition is not satisfied, you can use
*stopifnot*. This allow you to catch errors that can be anticipated.

Here's an example of building a robust square root function using
*stop* and *warning*. Note you could use `stopifnot(is.numeric(x))` in
place of one of the checks here.

```{r, eval=FALSE}
mysqrt <- function(x) {
	if (is.list(x)) {
		warning("x is a list; converting to a vector")
		x <- unlist(x)
	}
	if (!is.numeric(x)) {
		stop("What is the square root of 'bob'?")
        ## alternatively: stopifnot(is.numeric(x))
	} else {
		if (any(x < 0)) {
			warning("mysqrt: found negative values; proceeding anyway")
			x[x >= 0] <- (x[x >= 0])^(1/2)
			x[x < 0] <- NaN
			return(x)
		} else return(x^(1/2))
	}
}
mysqrt(c(1, 2, 3))
mysqrt(c(5, -7))
mysqrt(c("asdf", "sdf"))
mysqrt(list(5, 3, "ab"))
```

##  Catch run-time errors with  `try` statements

Also, sometimes a function you call will fail, but you want to
continue execution. For example, suppose you are doing a stratified
analysis in which you take subsets of your data based on some
categorical variable and fit a statistical model for each value of the
categorical variable. If some of the subsets have no or very few
observations, the statistical model fitting might fail. To do this,
you might be using a for loop or *lapply*. You want your code to
continue and fit the model for the rest of the cases even if one (or
more) of the cases cannot be fit.  You can wrap the function call that
may fail within the `try` function (or `tryCatch`) and then your code
won't stop, even when an error occurs. Here's a toy example.

```{r}
set.seed(0)
nCats <- 30
n <- 100
y <- rnorm(n)
x <- rnorm(n)
cats <- sample(1:nCats, n, replace = TRUE)
data <- data.frame(y, x, cats)

params <- matrix(NA, nr = nCats, nc = 2)
for(i in 1:nCats){
    sub <- data[data$cats == i, ]
    fit <- try(
        lm(y ~ x, data = sub) )
    if(!is(fit, "try-error"))
        params[i, ] = fit$coef
}
params
```

The seventh stratum had no observations, so that call to *lm* failed,
but the loop continued because we 'caught' the error with *try*. In
this example, we could have checked the sample size for the subset
before doing the regression, but in other contexts, we may not have an
easy way to check in advance whether the function call will fail.
